---
title: "DataSUS Tabnet data organisation"
author: "Raquel Martins Lana"
date: "2026-06-08"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

Warning: English version is not finished.

# 🌎 Course: Impacts of Climate Change on Health (PPGICS/Fiocruz)

> 🎯 **Learning Objectives**
>
> -   **Understand** the workflow of extracting and treating raw data
>     originating from DataSUS Tabnet.
> -   **Master** table manipulation and reformatting techniques
>     (horizontal and vertical formats).
> -   **Integrate** epidemiological databases with demographic data from
>     IBGE to calculate health indicators.
> -   **Build and interpret** graphical representations of time series
>     for the incidence of vector-borne diseases (VBDs).

> ## 🦟 Vector-Borne Diseases (VBD)
>
> #### 📊 Practical example: data on probable dengue cases (DataSUS Tabnet)
>
> ##### 🧑‍🏫 Lecturers: **Hermano Albuquerque, Izabel Reis and Raquel Lana (Lectures on 27/05/2026 and 03/06/2026)**

------------------------------------------------------------------------

### 📚 Epidemiological and climate context

The spatiotemporal distribution of Vector-Borne Diseases (VBDs), such as
dengue, Zika, and chikungunya, is strongly influenced by climate
variables (temperature, precipitation, and humidity). The global
increase in mean temperature and shifts in rainfall regimes expand the
ecological niches of vectors (such as the *Aedes aegypti* mosquito),
accelerating their reproductive cycle and shortening the extrinsic
incubation period of the virus. Analysing historical incidence data is
fundamental to anticipating outbreaks and structuring adaptation plans
for the public health system in the face of global environmental
changes.

------------------------------------------------------------------------

#### 📁 The Raw Data

The analytical data were extracted directly from [DataSUS
Tabnet](https://datasus.saude.gov.br/informacoes-de-saude-tabnet/),
comprising **probable dengue cases** aggregated by municipality of
residence (rows) and months of onset of the first symptoms (columns) for
the temporal window from 2020 to 2024. Each course working group
downloaded the file corresponding to a specific Brazilian state.

The main objective of this *script* is to carry out the **Data Cleaning
and Organisation** process, consolidating a unified table adapted to the
vertical (long) format, which is the ideal standard for advanced
modelling and graphical visualisations in R. The workflow was planned in
a semi-automated manner to mitigate repetitive tasks without obscuring
the fundamental concepts of the pedagogical learning process.

------------------------------------------------------------------------

#### ⚙️ Environment Configuration

In this stage, we prepare the computational operational environment,
verifying, installing (if necessary), and loading the essential
libraries for the methodological reproducibility of the protocol.

The packages used are:

`tidyverse`: a collection of packages designed for data science
(includes `ggplot2`, `dplyr`, `tidyr`, `readr`, among others).

`here`: a structural tool that uses relative paths based on the project
root, eliminating code breakage caused by absolute local directories.

`sf`: for working with geospatial data (Simple Features).

`geobr`: automatic *download* of official data and maps of geographic
boundary polygons from IBGE directly into R.

```{r Loading packages}
# 📦 List of required packages
packages <- c("tidyverse", "here", "sf", "geobr") 

# ⬇️ Install packages that are not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages], 
                   repos = "[https://cloud.r-project.org](https://cloud.r-project.org)")
}

# 🚀 Silent loading into operational memory
suppressPackageStartupMessages(
  invisible(lapply(packages, library, character.only = TRUE))
)

# 🚫 Avoid global scientific notation in the R session
# ⚙️ 'scipen' (scientific penalty) defines a "penalty" so R does not use scientific notation (such as 1e+05).
# 🔢 The default R value is 0. By changing it to 999, you force R to prefer fixed decimal notation (e.g. 100000).
options(scipen = 999)

# 🔄 To revert to the default R scientific notation, uncomment the line below:
# options(scipen = 0)

cat("✅ Computational environment configured successfully!\\n")
```

### 📥 Loading Data and First Cleaning Stage

⚠️ **Methodological alert:** when exporting reports in `.csv` format
from Tabnet, the system generates explanatory textual metadata in the
first and last lines of the file. Attempting to read these files
directly without treating these lines will result in a critical
column-typing failure (all columns will be converted to text due to the
informative header strings). To clarify, a string in its literal
translation is a "string of characters" — here it means any text inside
a computer code. A variable that is a string is always enclosed in
single or double quotation marks.

💡 **Tip**: use the `readLines()` function before loading a new,
unfamiliar database. It allows you to visually inspect the character
structure, identify the exact separator character (`;` or `,`), and
count how many rows should be ignored at the beginning (`skip`).

```{r Looking at initial lines and separators}
readLines(
  here("dados", "minas_gerais_2020.csv"),
  n = 15)

readLines(
  here("dados", "minas_gerais_2021.csv"),
  n = 15)

readLines(
  here("dados", "minas_gerais_2022.csv"),
  n = 15)

readLines(
  here("dados", "minas_gerais_2023.csv"),
  n = 15)

readLines(
  here("dados", "minas_gerais_2024.csv"),
  n = 15)
```

With this, we can observe that the first 3 lines need to be excluded, as
they are metadata rows. We also verified the separators.

This code is normally used when importing text files (`.csv`, `.txt`,
etc.) to ensure the correct interpretation of characters and data types.
The first command creates a regional configuration using the `locale`
function. The parameters are:

-   `encoding` = "latin1": states that the file uses `Latin-1`
    (ISO-8859-1) encoding, very common in files produced by older
    Brazilian systems.

-   `decimal_mark` = "," : the decimal separator used in numbers for
    this example, meaning 12,5 will be interpreted as 12.5.

-   `grouping_mark` = "." : the thousands separator, meaning 1.234 will
    be interpreted as 1234.

-   `asciify` = FALSE: preserves accented characters (á, é, ç, ã, etc.)
    without converting them to ASCII format.

The second command defines a column specification for data reading:

-   `.default` = "c": Forces the initial import of all columns as
    characters, preventing empty columns or columns encoded with special
    characters from breaking the data import.

```{r Loading the data and deleting the first few rows}
# Regional configuration adapted for Brazilian public data
pt_br <- locale(
  encoding = "latin1",
  decimal_mark = ",",
  grouping_mark = ".",
  asciify = FALSE
)

colunas <- cols(.default = "c")

# 📥 Individual import of annual matrices
d20 <- read_csv2(here("dados/minas_gerais_2020.csv"), 
                 locale = pt_br, 
                 col_types = colunas, 
                 skip = 3)

d21 <- read_csv2(here("dados/minas_gerais_2021.csv"), 
                 locale = pt_br, 
                 col_types = colunas, 
                 skip = 3)

d22 <- read_csv2(here("dados/minas_gerais_2022.csv"), 
                 locale = pt_br, 
                 col_types = colunas, 
                 skip = 3)

d23 <- read_csv2(here("dados/minas_gerais_2023.csv"), 
                 locale = pt_br, 
                 col_types = colunas, 
                 skip = 3)

d24 <- read_csv2(here("dados/minas_gerais_2024.csv"), 
                 locale = pt_br, 
                 col_types = colunas, 
                 skip = 3)
```

⚠️ **Methodological Alert**: *Warning* messages occur because the
explanatory footer of the file exceeds the tabular limits configured for
the columns. To resolve this, we isolate exclusively the rows that
represent actual municipalities, excluding the entire footer, and we
also remove the last column of horizontal totals, keeping only the
strict monthly distribution. The message should look like this, do not
be alarmed!

```{=html}
Warning: One or more parsing issues, call problems() on your data frame for 
details, e.g.: dat <- vroom(...) problems(dat)
```

The best option now is to open the file and see which rows we need to
exclude at the end of each year. Notice that up to row 854 of `d20`, we
have the expected pattern. From the "Total" row onwards, we must exclude
it, as well as the last column named "Total", which is column 14. Even
if all years have the same number of rows, it is important to open them
one by one and check. This was only possible because when downloading
the data, it was requested that zero-count municipalities were also
included! Do you remember?

💡 **Tip**: as you can see from the way the data were downloaded, there
is no difference in the quantity of municipalities in Minas Gerais. This
probably means that over these years, there was no change in the number
of municipalities in the state.

```{r Excluding lines and columns}
# ✂️ Manual cleaning based on the physical limit of actual records in each table
d20 <- d20[1:854, -14]
d21 <- d21[1:854, -14]
d22 <- d22[1:854, -14]
d23 <- d23[1:854, -14]
d24 <- d24[1:854, -14]

# 📝 Checking
tail(d20)
tail(d21)
tail(d22)
tail(d23)
tail(d24)
```

## 🛠️ Second Stage: Structural Column Formatting

The original variable regarding the municipality of residence stores two
distinct units of information within the same string: the municipal
geocode (the first 6 digits of the column) and the literal name of the
municipality. To enable future cross-referencing with demographic or
cartographic databases, we need to separate this column into two new
columns.

Additionally, in the case record information per month, we replace the
hyphen character "-" (used by the system to signal the absence of
records) with the strictly numerical value 0.

To achieve this, the `format_tabnet()` function was created. To use it,
you only need to provide the table or object in question.

⚠️ **Methodological alert**: the IBGE municipal geocode has 7 digits,
whereas DataSUS uses 6. Therefore, at times, it will be necessary to
exclude the seventh digit from data originating from IBGE. We will not
encounter this issue in this routine.

```{r Reformating lines and columns}
# ⚙️ Declaration of the modular structural standardisation function
# Run the function below once
format_tabnet <- function(df){

  df %>%
    mutate(
      geocode = str_sub(`Município de residência`, 1, 6),
      municipio = str_sub(`Município de residência`, 7)
    ) %>%
    select(-`Município de residência`) %>%
    mutate(
      across(
        Jan:Dez,
        ~ as.numeric(str_replace_all(.x, "-", "0"))
      )
    ) %>%
    mutate(
      municipio = str_trim(
        str_remove_all(municipio, '"')
      )
    )
}

# ⚡ Application of the automated cleaning routine to the annual databases
d20 <- format_tabnet(d20)
d21 <- format_tabnet(d21)
d22 <- format_tabnet(d22)
d23 <- format_tabnet(d23)
d24 <- format_tabnet(d24)
```

Before proceeding to the next stage, open one of the years and see how
it turned out!

## 📐 Verticalisation of Data Matrices

To conduct longitudinal analyses, we structure the data into a long
(vertical) format, in which each row maps a unique spatiotemporal
observation (Municipality, Month, Year, and Cases).

```{r Verticalise data}
# 🔁 Function for horizontal-to-vertical structural pivoting
# Run the function to verticalise the data and add the year variable
vertical_dados <- function(df, ano) {
  df %>%
    gather(meses, casos, -c(geocode, municipio)) %>%
    mutate(ano = ano)
}

# ⚡ Processing and chronological indexing per specific year
# Apply the function to each year
d20 <- vertical_dados(d20, 2020)
d21 <- vertical_dados(d21, 2021)
d22 <- vertical_dados(d22, 2022)
d23 <- vertical_dados(d23, 2023)
d24 <- vertical_dados(d24, 2024)
```

💡 **Tip**: compare the difference using the `head(d20)` command in your
R console to visually inspect how the table was restructured in an
optimised way for data analysis.

Once the columns are normalised, we can perform a full row binding of
all annual tables into a single data object.

⚠️ **Methodological alert**: for this to work correctly, you must first
ensure that the columns are of the same data type. Since we loaded the
data requesting all variables as characters, all columns across all
years should share the same type. However, if you want to be certain,
execute the `str(d20)` command in your R console, and so on for all
years. If it is necessary to change a column type, use the example code
below, adapting it to your case.

```{r Joint years}
# Changing a variable type - example
# dexemplo <- dexemplo %>% 
#  mutate(var1 = as.character(var1))
# To change the type of more than one variable at the same time
# dexemplo <- dexemplo %>% 
#  mutate(across(c(var1, var2, var3), as.character))


# 🧲 Unified consolidation by row binding
d <- bind_rows(d20,
               d21,
               d22,
               d23,
               d24)
```

##### 📅 Building the Standard Date Variable

To structure and parameterise the chronological scales in vector
time-series plots, we convert the year and month strings into a native R
`Date` class object, systematically indexing all observations to the
first day of each month.

```{r date variable}
# 🗺️ Mapping text to integers and classic chronological formatting
d <- d %>%
    mutate(
      data = as.Date(paste0(ano, "-", 
                            match(meses,
                            c("Jan","Fev","Mar","Abr","Mai","Jun","Jul",
                              "Ago","Set","Out","Nov","Dez")),
                            "-01"))) %>% 
  select(muni_code = geocode, muni_name = municipio, data, meses, ano, casos)
```

Let's check the variables?

```{r Checking}
# 🔍 Identification of residual artifacts and registry noise
unique(d$muni_code) 
# ⚠️ Methodological Alert: The dummy code "MUNICI" represents "Unknown Municipality"
# This marker occurs when the system cannot identify the patient's city of residence.

# 🧼 Systematic exclusion of noise to ensure spatial integrity
d %>% filter(muni_code == "MUNICI")

d <- d %>% 
  filter(muni_code != "MUNICI") # 853

# ✅ Final validation of consistency and chronological completeness
unique(d$muni_code)
# How many unique months? Is any missing?
unique(d$meses)
# And the years?
unique(d$ano)

# How many total months?
length(unique(d$data)) # Expected: 60 (12 months x 5 sampling years)
```

#### 🔄 Matrix Expansion for Municipalities Without Active Notification

⚠️ **Methodological alert:** this stage may be necessary in certain
cases, see below:

1.  This stage applies only if there are differences in the
    municipalities that appear in each year.

2.  If a municipality did not register notifications in a given year, it
    remains absent from the `csv` file generated by Tabnet, unless
    zero-counts were requested. If we compute averages and statistics
    directly from this omitted table, we will generate a severe
    epidemiological distortion (overestimating the real incidence). We
    need to force the inclusion of all 853 municipalities of Minas
    Gerais across all 60 historical months of the study, inserting the
    value zero into the actual gaps.

The resulting expanded database will contain exactly

$853 \text{ municipalities} \times 12 \text{ months} \times 5 \text{ years} = 51,180 \text{ rows}$.

**If this is not your case, skip to the next section.**

```{r Expand grid}
# 🔢 Calculation of the theoretical baseline size for validation
total_linhas <- 853 * 12 * 5
cat("Expected total lines:", total_linhas, "\\n")

# 📅 Extraction of the temporal reference calendar without duplicates
calendario <- d %>%
  distinct(data, ano, meses)

# 🗺️ Isolation of the unified georeferenced municipality register
municipios <- d %>%
  distinct(muni_code, muni_name)

# 🧬 Combinatorial expansion matching each municipality to all possible temporal units
dexp <- merge(municipios, calendario, all = TRUE)

# 🔀 Structural merging and controlled imputation of missing data (zeros)
dcompleto <- dexp %>%
  left_join(
    d,
    by = c("muni_code", "muni_name", "data", "ano", "meses")
  ) %>%
  mutate(casos = coalesce(casos, 0))

# ✅ Checking
unique(dcompleto$muni_code) 

# How many unique months? Is any missing?
unique(dcompleto$meses)
# And the years?
unique(dcompleto$ano)

# How many total months?
length(unique(dcompleto$data)) # 60: 12 months x 5 years

```

### 👥 Integration of Demographic Data and Population Estimates

To convert the absolute number of cases into an epidemiological metric
that is comparable across different territories, we need to weight the
records by the size of the population exposed to the risk. We will use
the consolidated historical estimates from the Census and TCU, organised
and publicly shared on the [GitHub repository of researcher Leonardo
Bastos](https://github.com/lsbastos/popBR_mun) from Fiocruz.

> **`.csv`:** a universal format, readable by any software (Excel,
> Python, QGIS, etc.). Ideal for sharing data with other users or using
> them outside R. The disadvantage is that column types (dates, factors,
> etc.) are not automatically preserved upon reading — they need to be
> reconverted.

```{r Population}
# ⬇️ Direct import via URL channel of the long-format population data repository
pop <- read.csv("[https://raw.githubusercontent.com/lsbastos/popBR_mun/refs/heads/master/popBR2000-2024.long.csv](https://raw.githubusercontent.com/lsbastos/popBR_mun/refs/heads/master/popBR2000-2024.long.csv)")

# 📍 Spatial subset restricted to the state of Minas Gerais (IBGE Prefix: 31)
pop_mg <- pop %>%
  filter(
    str_starts(as.character(MUNCOD), "31"),
    Ano %in% 2020:2024
  )

unique(pop_mg$Ano)
unique(pop_mg$MUNCOD)

# 💡 Tip for Students: Save this population file to your local hard drive
# to avoid redundant and time-consuming web requests on every new execution of the script.
# write.csv(pop_mg, "dados/pop_mg.csv", row.names = F) # -----> EDIT the file name

# If you need to reload these data, if already saved, run:
# pop_mg <- read_csv("dados/pop_mg.csv") # -----> EDIT the file name
```

The content of the table is:

-   `MUNCOD` - municipality code

-   `MUN` - municipality name

-   `Ano` - year

-   `POP` - population per municipality/year (Census for the years 2010
    and 2022, and projected via TCU for the years 2011 to 2021).

#### 🔗 Database Fusion (Cases $\times$ Population) 

To easily add the population to the cases table, let's edit some column
names to facilitate the table join. We also format the variable types to
make the logical key or linking fields compatible.

```{r Joint tables}
# 🔧 Adjustment of nomenclature and column typing for relational joining
pop_mg <- pop_mg %>% 
  mutate(muni_code = as.character(MUNCOD)) %>% 
  rename(ano = Ano) %>% 
  select(-c(MUNCOD))

# 🖇️ Fusion of tables and cleaning of redundant duplicate columns
d <- left_join(d, pop_mg) %>% 
  select(-MUN) %>% 
  rename(pop = POP)
```

#### 🧮 Calculation of the Incidence Rate

The incidence rate expresses the risk of new cases occurring in the
population over a specific period, traditionally measured for dengue
using a multiplier base of 100,000 inhabitants.

\$\$\\text{Incidence} = \\left( \\frac{\\text{Active
Cases}}{\\text{Resident Population}} \\right) \\times 100,000\$\$

```{r Incidence}
# 🖩 Analytical calculation of the population incidence rate indicator
d$incidencia <- d$casos / d$pop * 100000  

head(d)
```

#### 💾 Saving the Consolidated Database

If you have not saved the database yet, uncomment and execute the line
below:

```{r Saving table}
# 💾 Physical saving of the final file in the results folder
# write.csv(d, "resultados/dengue_mensal_minas_gerais_2020-2024.csv", row.names = F)
```

### 📈 Data Visualisation and Graphical Analysis (Time Series)

With the treated database, we begin the visual analysis phase of
epidemiological patterns across different analysis scales.

📍 **Municipal Scenario 1**: historical evolution of cases in Abaeté
(MG)

```{r}
# 📉 Plotting the evolution of absolute cases in Abaeté
# 📦 1. Stores the plot in the object 'plot_abaete', making it easier to save the figure
plot_abaete <- d %>%
  filter(muni_code == 310020) %>%
  ggplot(aes(x = data, y = casos)) +
  geom_line(linewidth = 0.8, color = "blue") +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b/%Y"
  ) +
  labs(
    title = "Monthly time series of dengue cases",
    subtitle = "Abaeté (MG), 2020–2024",
    x = NULL,
    y = "Cases"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5),
    axis.ticks = element_line(color = "black"),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 8),
    plot.title = element_text(size = 10, face = "bold")
  )


# 💾 2. Command to save the plot above as a png
ggsave(
  plot = plot_abaete,                           # 🎯 Indicates which plot to save
  filename = "resultados/grafico_abaete_dengue.png", # 📂 Path and file name (change folder if necessary)
  width = 7,                                    # 📏 Width in inches
  height = 5,                                   # 📐 Height in inches
  dpi = 300                                     # 🔍 High resolution (ideal for articles/reports)
)


# 💾 3. Command to save the plot above as a pdf
ggsave(
  plot = plot_abaete,                      # 🎯 Indicates which plot to save
  filename = "grafico_abaete_dengue.pdf",  # 📄 Saving in PDF format (vectorial, does not lose quality)
  width = 18,                              # 📏 Dimensions in centimetres if changing 'units'
  height = 12,
  units = "cm"                             # ⚙️ Defines the measurement unit (can be "in", "cm" or "mm")
)
```

💡 Technical Recommendations:

-   **Dimensions** (`width` and `height`)**:** setting the size manually
    prevents R from saving the plot based on your current RStudio screen
    size, which usually distorts text and legends.

-   **Formats:** use `.png` for digital presentations or Word/Markdown,
    and `.pdf` if you need to submit for scientific publication
    requiring high graphical quality. There are other possibilities too,
    such as `.tiff`, `.jpg`, `.svg`.

📐 Change the dimensions and see what happens! 🔄 💾 Try saving the next
figures using the example above!!! 📉

📍 **Municipal Scenario 2**: incidence rate in Belo Horizonte (MG)

```{r}
# 📉 Plotting the standardised incidence rate in the state capital
d %>%
  filter(muni_code == 310620) %>%
  ggplot(aes(x = data, y = incidencia)) +
  geom_line(linewidth = 0.8, color = "red") +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b/%Y"
  ) +
  labs(
    title = "Dengue incidence in Belo Horizonte (MG)",
    x = NULL,
    y = "Incidence per 100,000 inhabitants"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5),
    axis.ticks = element_line(color = "black"),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 10, face = "bold")
  )
```

📈 **Evidence-based interpretation**: the visual scale generated by the
epidemiological year of 2024 is so massive that it causes a sharp
flattening effect on the previous years (2020 to 2023). This indicates
that the magnitude of the 2024 outbreak was atypical compared to the
recent historical series of Belo Horizonte.

💡 **Tip**: when an extreme event flattens the history of your series,
temporarily remove the atypical year to clearly investigate the seasonal
dynamics and background behaviour of previous periods.

```{r}
# 📉 Analytical zoom excluding the large 2024 epidemic
d %>%
  filter(muni_code == 310620, ano != 2024) %>%
  ggplot(aes(x = data, y = incidencia)) +
  geom_line(linewidth = 0.8, color = "red") +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b/%Y"
  ) +
  labs(
    title = "Dengue incidence in Belo Horizonte (MG)",
    x = NULL,
    y = "Incidence per 100,000 inhabitants"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5),
    axis.ticks = element_line(color = "black"),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 10, face = "bold")
  )
```

📈 **Evidence-based interpretation (zoom without 2024)**: without the
extreme epidemic year of 2024, the cyclical and highly seasonal
behaviour of dengue in the capital becomes evident. The peaks invariably
occur in the first half of each year (coinciding with the warmest and
rainiest months of the transition from summer to autumn). There is also
a clear interannual variation: the year 2021 showed a visibly reduced
transmission, contrasting with the epidemic year of 2023. Perhaps it is
still worth removing 2023 to visualise the rest, give it a try!!!!

#### 🗺️ Macro-regional Scenario: Consolidated Aggregation of the State of Minas Gerais

To generate a synthesis of the health issue's behaviour across the
territory, we aggregate the numerical values of all 853 municipalities,
estimating a unified monthly indicator for the entire state of Minas
Gerais.

```{r}
# 🏗️ Spatiotemporal grouping and macro-epidemiological summarisation
dmg <- d %>%
  group_by(data) %>% 
  summarise(casos = sum(casos),
            pop = sum(pop)) %>% 
  mutate(incidencia = casos / pop * 100000)
```

See by state:

```{r}
# 📉 Longitudinal monitoring of aggregated state cases
dmg %>%   
  ggplot(aes(x = data, y = casos)) +
  geom_line(linewidth = 0.8) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b/%Y"
  ) +
  labs(
    title = "Dengue cases in Minas Gerais",
    x = NULL,
    y = "Cases"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5),
    axis.ticks = element_line(color = "black"),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 10, face = "bold")
  )
```

Now, the incidence!

```{r}
# 📉 Consolidated incidence rate for the state of Minas Gerais
dmg %>%   
  ggplot(aes(x = data, y = incidencia)) +
  geom_line(linewidth = 0.8) +
  scale_x_date(
    date_breaks = "6 months",
    date_labels = "%b/%Y"
  ) +
  labs(
    title = "Dengue incidence in Minas Gerais",
    x = NULL,
    y = "Incidence per 100,000 inhabitants"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", linewidth = 0.5),
    axis.ticks = element_line(color = "black"),
    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8),
    axis.title.y = element_text(size = 9),
    plot.title = element_text(size = 10, face = "bold")
  )
```

📈 **Evidence-based interpretation (state scenario)**: the aggregated
dynamics for the state reflect the same pattern observed in Belo
Horizonte, confirming that the 2024 epidemic was a widespread phenomenon
on a regional scale rather than a localised outbreak. Incidence peaks
occur synchronously in the first half of each year. Continuous
monitoring of these curves is of great importance for planning and
emergency resource allocation by state public health authorities.

#### 🗺️ Map by Municipality

To produce a map of cases or incidence for the state by municipality, it
is necessary to obtain the geographic boundary polygons. The function
below allows you to do this automatically via the `geobr` package. The
data were downloaded for the year 2020. To avoid repeated downloads on
every execution of the script, the code to save and load the map locally
is available below.

```{r}
# 🗺️ Downloading the map of MG municipalities for the year 2020
# Note: the IBGE standard comes with 7 digits (we will remove the last check digit for merging).
mapa <- read_municipality(code_muni = "MG", year = 2020) # ✏️ EDIT the state and year if necessary

# 💾 Optimisation: save the map locally so you do not need to download it from the internet every time
# saveRDS(mapa, here::here("dados/mapa_mg.rds"), compress = "xz")

# 📂 To load the saved map from your computer in future executions, uncomment the line below:
# mapa <- readRDS(here::here("dados/mapa_mg.rds"))

# 🔍 Inspecting the map column structure
str(mapa)

# ⚙️ Harmonising the linking variable: creating the 6-digit code for the municipality.
# 'str_sub' automatically converts the number to a character (string) before slicing, saving lines of code!
mapa <- mapa %>% 
  mutate(muni_code = str_sub(code_muni, 1, 6))
```

🧩 Now it is time to join the map with the dengue data: 🗺️📊

```{r}
# 📊 Filtering epidemiological data for the year 2020 and consolidating by municipality
d20_anual <- d %>% 
  filter(ano == 2020) %>% 
  group_by(muni_code) %>% 
  summarise(casos = sum(casos, na.rm = TRUE)) # 'na.rm = TRUE' prevents missing values from zeroing the sum

# 🔀 Merging spatial data (map) with health data (cases) using the common code
mapa_casos20 <- left_join(mapa, d20_anual, by = "muni_code")
```

##### 🎉 We have a beautiful map! Try creating the incidence one 📊🦟

Observe in detail what each part of the code does!

```{r}
# 🎨 Building the map with ggplot2
m20 <- ggplot(data = mapa_casos20) +
  # 🟢 Draws the municipality geometry (sf) and fills color based on the number of cases
  geom_sf(aes(fill = casos), color = "white", size = 0.1) +
  
  # 🌈 "Viridis" color palette (highly recommended: great for color-blindness and scientific printing)
  scale_fill_viridis_c(
    option = "plasma", 
    name = "No. of Cases",
    na.value = "gray90" # 🔘 Light grey color for municipalities without data records
  ) +
  
  # 📝 Titles, subtitles and footnotes
  labs(
    title = "Case distribution by municipality",
    subtitle = "Minas Gerais, 2020",
    caption = "Source: DataSUS and IBGE / Author's own elaboration"
  ) +
  
  # 🧼 "void" theme (clean): removes axes, coordinates and grids, highlighting only the map
  theme_void() + 
  
  # 📐 Fine-tuning typography and element positioning
  theme(
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, face = "italic"),
    legend.position = "right"
  )

# 💾 Saving the case distribution map
ggsave(
  plot = m20,                                # 🎯 Specifies the stored map object
  filename = "resultados/mapa_mg_casos_2020.png", # 📂 Output path and file name
  width = 8,                                 # 📏 Width calibrated for the geometric format of MG
  height = 7,                                # 📐 Height balanced to maintain spatial proportion
  dpi = 300                                  # 🔍 High resolution for reports and publications
)
```

💡 **Golden tip for maps:** maps based on `sf` (Spatial Features)
objects are sensitive to scale distortion. If you notice that the map of
Minas Gerais looks too "stretched" or "flattened", slightly change the
`width` or `height` values until you find the perfect fit for your
layout.

##### 💾 Exporting in multiple formats 

Uncomment the code if you wish to save in one of these formats. For the
last one, the specified package is required.

```{r}
# 🖼️ JPG (Ideal for presentations and quick sharing)
# ggsave(
#   filename = "resultados/mapa_mg_casos_2020.jpg",
#   plot = m20, 
#   width = 8, 
#   height = 7, 
#   dpi = 300
# )

# 📄 PDF (Vector format standard for academic reports)
# ggsave(
#   filename = "resultados/mapa_mg_casos_2020.pdf",
#   plot = m20, 
#   width = 8, 
#   height = 7
# )

# 🖨️ TIFF (Lossless high quality, required by scientific journals)
# ggsave(
#   filename = "resultados/mapa_mg_casos_2020.tiff",
#   plot = m20, 
#   width = 8, 
#   height = 7, 
#   dpi = 300,
#   compression = "lzw" # Reduces file size without losing quality
# )

# 🌐 SVG (Native vector for web and editing in Inkscape/Illustrator)
# Note: Requires the 'svglite' package installed (install.packages("svglite"))
# ggsave(
#   filename = "resultados/mapa_mg_casos_2020.svg",
#   plot = m20, 
#   width = 8, 
#   height = 7
# )
```

#### ✅ Final Pedagogical Conclusion

In this practical laboratory activity, we consolidated a complete
processing workflow for public health data: from raw DataSUS Tabnet CSV
files to a clean database ready for use in scientific analyses. We
learned how to bypass common inconsistencies such as metadata rows,
regional character encodings, and structural omissions of municipalities
without records.

🔄 Transforming the data into the long format and the subsequent
integration with IBGE population projections allowed us to move from the
analysis of absolute numbers to standardised epidemiological indicators.

🗺️ Furthermore, we advanced to spatial analysis, integrating our data
with IBGE polygons via the `geobr` package. We succeeded in generating
an absolute case distribution map for the state and set up the
scientific challenge of replicating this same visual logic for the
incidence rate per 100,000 inhabitants — a fundamental metric to compare
municipalities of different sizes without population bias.

🌍 Mastering this entire methodological workflow — which unites data
treatment, rate calculation, and cartography — constitutes the core
foundation for studies using statistical and predictive modelling
applied to the impacts of global climate changes on human health.
